---
title: Data Visualization
---

<!--
Possible prerequisites
- Quarto intro
- R data wrangling
-->

:::{.callout-learning}
After completing this session, you will be able to:

- Describe the four essential components of a `ggplot2::ggplot()` call
- Use `theme` and other functions to customize appealing data visualizations
- Use `facet_wrap()` to create multiple related plots in one figure
- Introduce the `leaflet` and `DT` package to create interactive maps and tables respectively
:::

## Overview

`ggplot2` is a popular package for visualizing data in R.
From the [home page](http://ggplot2.tidyverse.org/):

> `ggplot2` is a system for declaratively creating graphics, based on The Grammar of Graphics. You provide the data, tell `ggplot2` how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details.

The goal of this lesson is to explain the fundamentals of how `ggplot2` work, introduce useful functions for customizing your plots, and inspire you to go and explore this awesome resource for visualizing your data.

::: callout-note

## `ggplot2` vs base graphics in R vs others
There are many different ways to plot your data in R.  All of them can be used to make functional plots! However, `ggplot2` uses a reasonably intuitive syntax to build up a plot in layers and customize it in many different ways. It also has a lot of built in themes and color palettes that can make your plots look nice with just a few lines of code.

Base R graphics  (`plot()`, `hist()`, etc) can be helpful for simple, quick-and-dirty plots, e.g., for data exploration, but it can be hard to decipher and customize. `ggplot2` can be used for almost everything else, including publication-ready graphics.

:::

## Setup

Make sure you’re in the right project (`training_{USERNAME}`). Use Git Pull to check for any changes. Then, create a new Quarto document, delete the default text below the YAML header, and save this document.  

:::{.callout-tip collapse=true}
## Quarto header options?

Consider adding Quarto options to the YAML header to control the output:
```yaml
format:
  html:
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  warning: false
  message: false
```
:::

### Load Packages

Load the packages we'll need. They're grouped by purpose to make it easier to remember what each one does.

```{r load_packages}
#| message: false
#| warning: false

### Data wrangling
library(readr)
library(dplyr)
library(tidyr)
library(forcats)   ### tools for working with factors (e.g., reordering)
library(janitor)   ### clean column names

### Visualization
library(ggplot2)
library(scales)    ### axis formatting helpers
library(viridis)   ### colorblind-friendly palettes

### Interactive output
library(leaflet)
library(DT)
```

### Load the Data

Download the data from the [EDI Data Repository](https://portal.edirepository.org/nis/mapbrowse?packageid=edi.587.1): scroll to *Resources*, click the "Socioecological monitoring data" download button, and save to your `data/` folder. Then read it in:

```{r}
#| echo: false
#| eval: true

### NOTE THIS CHUNK IS HIDDEN (echo: false) AND EVALUATED (eval: true)
### FOR THE PURPOSES OF RENDERING THE LESSON, WITH THE EXTRA FOLDER IN PATH.
### THE FOLLOWING CHUNK (echo: true and eval: false) IS FOR WHERE USERS SHOULD
### SAVE FILES.

delta_visits_raw <- read_csv(file.path("data", "r_data_visualization", 
                                       "Socioecological_monitoring_data.csv"),
                             show_col_types = FALSE)
```

```{r}
#| echo: true
#| eval: false

# read in data from the data directory after manually downloading data 
delta_visits_raw <- read_csv("data/Socioecological_monitoring_data.csv")

```

### Learn About the Data

For this session we are going to be working with data on **Socioecological Monitoring on the Sacramento-San Joaquin Delta**. Check out [the documentation](https://portal.edirepository.org/nis/mapbrowse?packageid=edi.587.1): read the abstract, explore the metadata, check the intellectual rights, etc. 

### Explore the Data

Finally, let's explore the data we just read into our working environment. 

```r
### Check out column names
colnames(delta_visits_raw)

### Peek at each column and class
glimpse(delta_visits_raw)

### From when to when
range(delta_visits_raw$Date)

### Which time of day?
unique(delta_visits_raw$Time_of_Day)
```


## Prepare the Data

We nearly always need to do some wrangling before we can analyze and plot our data.  Here we will wrangle our data into a tidy format that is `ggplot2`-friendly.

### Clean Column Names

`janitor::clean_names()` converts all column names to consistent `snake_case` in a single line:

```{r}
colnames(delta_visits_raw)[1:6]

delta_visits <- delta_visits_raw %>%
  janitor::clean_names()

colnames(delta_visits)[1:6]
```

### Reshape to Long (Tidy) Format

Recall the [tidy data principles](https://r4ds.hadley.nz/data-tidy.html#fig-tidy-structure){.external target="_blank"}.

1. Each variable is a column; each column is a variable.
2. Each observation is a row; each row is an observation.
3. Each value is a cell; each cell is a single value.

Which (if any) of these principles are violated in our data frame `delta_visits`?

`ggplot2` works best with **tidy data** — one row per observation. Our visitor-type counts are spread across multiple columns, so we pivot them (using `tidyr::pivot_longer()`) into a single `visitor_type` column:

```{r}
visits_long <- delta_visits %>%
  tidyr::pivot_longer(
    cols = c(sm_boat, med_boat, lrg_boat, bank_angler, scientist, cars),
    names_to = "visitor_type",
    values_to = "quantity"
  ) %>%
  ### shorten the long name:
  dplyr::rename(restore_loc = eco_restore_approximate_location) %>%
  ### drop the notes column:
  dplyr::select(-notes)

head(visits_long)
```



::: callout-exercise

Using the `visits_long` dataframe, calculate the daily visits by `restore_loc`, `date`, and `visitor_type`.  Assign the result to a new object called `daily_visits_loc`.  Then inspect the result.

::::{.callout-answer}
 
```{r mean_esc}
daily_visits_loc <- visits_long %>%
    dplyr::group_by(restore_loc, date, visitor_type) %>% 
    dplyr::summarize(daily_visits = sum(quantity), .groups = 'drop')
    
head(daily_visits_loc)
```

Refresher:

* `group_by()` to calculate our results for the unique combinations of type of visit, restoration location proximity, and day.  
* `summarise()` to sum up the daily visit value for each of these groups. 
* `%>%` operator to pipe in the result of one command as an argument to the next one.

::::

:::

## Plot With `ggplot2`

### Essential Components

Every `ggplot2` call requires four essential components:

1. **`ggplot()`** opens the plot (note, the **package** is called `ggplot2` but the function is just `ggplot()`.)
2. **`data =`** defines the data frame to use
3. **`aes()`** maps variables to visual properties (axes, color, fill, etc.)
4. **`geom_*()`** defines the plot type (bar, point, line, box, etc.)

Layers are added with `+`.

```{r}
#| fig-align: "center"
#| fig-height: 3.5

ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits)) +
  geom_col()
```

::: callout-tip
`data =` and `aes()` can also live inside the `geom_*()` call — useful when layering multiple data sources!  But for simple plots, a good practice is to keep them in `ggplot()` for clarity.

**Variations on `data=` and `aes()`**

Options 1 - 3 produce the same output as above.  Option 4 flips the axes.  Examine the code and the outputs.

:::: panel-tabset
## Option 1 (original)

Data and mapping defined in `ggplot()` function:
```{r}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5

ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits)) +
    geom_col()
```

## Option 2

Data defined in `ggplot()` function; mapping defined in `geom_*()`:
```{r}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5
ggplot(data = daily_visits_loc) +
    geom_col(aes(x = restore_loc, 
                 y = daily_visits))
```

## Option 3

Data and mapping both defined in `geom_*()`
```{r}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5
ggplot() +
    geom_col(data = daily_visits_loc,
             aes(x = restore_loc, y = daily_visits))
```

## Option 4 (axes flipped)

Data and mapping both defined in `geom_*()`, but with x and y flipped.  Note the difference!
```{r}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5
ggplot() +
    geom_col(data = daily_visits_loc,
             aes(y = restore_loc, x = daily_visits))
```

:::: 
<!-- end tabset -->
:::
<!-- end tip -->


### Different `geom_*` Types

Having the basic structure with the essential components in mind, we can easily change the type of graph by updating the `geom_*()`.  First we will add a few lines to further process the data.

:::panel-tabset

## Boxplot

```{r boxplot_ex}
daily_visits_loc %>%
    separate(date, c("year", "month", "day"), sep = "-") %>%
    filter(daily_visits < 30,
           visitor_type %in% c("sm_boat", "med_boat", "lrg_boat")) %>%
    ggplot(aes(x = visitor_type, y = daily_visits)) +
    geom_boxplot()
```


**Violin plot**

```{r violin_plot}
daily_visits_loc %>%
    separate(date, c("year", "month", "day"), sep = "-") %>%
    filter(daily_visits < 30,
           visitor_type %in% c("sm_boat", "med_boat", "lrg_boat")) %>%
    ggplot(aes(x = visitor_type, y = daily_visits)) +
    geom_violin()
```


**Line and point**

```{r line_poin_plot}

daily_visits_loc %>%
    filter(restore_loc == "Decker Island",
           visitor_type == "med_boat") %>%
    ggplot(aes(x = date, y = daily_visits)) +
    geom_line() +
    geom_point()
```


### Customizing our plot

Let's go back to our base bar graph. What if we want our bars to be blue instead of gray? You might think we could run this:

```{r fill_blue}

ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits,
           fill = "blue")) +
    geom_col()
```

Why did that happen?

Notice that we tried to set the fill color of the plot inside the mapping aesthetic call. What we have done, behind the scenes, is create a column filled with the word "blue" in our data frame, and then mapped it to the `fill` aesthetic, which then chose the default fill color, salmon. 

What we really wanted to do was just change the color of the bars. If we want do do that, we can call the color option in the `geom_col()` function, **outside** of the mapping aesthetics function call.

```{r fill_blue_geom}

ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits)) +
    geom_col(fill = "blue")

```

What if we did want to map the color of the bars to a variable, such as `visitor_type`. `ggplot()` is really powerful because we can easily get this plot to visualize more aspects of our data. 

```{r}

ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits,
           fill = visitor_type)) +
    geom_col()

```

::: {.callout-tip}

### Keep in mind

- If you want to map a variable onto a graph aesthetic (e.g., point color should be based on a specific region), put it within `aes()`.

- If you want to update your plot base with a constant  (e.g. “Make ALL the points BLUE”), you can add the information directly to the relevant `geom_` layer outside the `aes()` call.

:::


#### Setting `ggplot` themes


We have successfully plotted our data. But, this is clearly not a nice plot. Let's work on making this plot look a bit nicer. We are going to:

- Add a title, subtitle and adjust labels using `labs()`
- Flip the x and y axis to make it a sideways column plot and make the labels easier to read
- Include a built in theme using `theme_bw()`


```{r theme_bw_plot}

ggplot(data = daily_visits_loc,
       aes(y = restore_loc, x = daily_visits, fill = visitor_type)) +
    geom_col() +
    labs(x = "Number of Visits",
         y = "Restoration Location",
         fill = "Type of Visitor",
         title = "Total Number of Visits to Delta Restoration Areas by visitor type",
         subtitle = "Sum of all visits during July 2017 and March 2018") +
    theme_bw()
    
    
```

You can see that the `theme_bw()` function changed a lot of the aspects of our plot! The background is white, the grid is a different color, etc. There are lots of other built in themes like this that come with the `ggplot2` package that help quickly set the look of the plot. Use the RStudio auto-complete `theme_` `<TAB>` to view a list of theme functions.


```{r}
#| code-summary: "Themes"
#| eval: false

## Useful baseline themes are
theme_minimal()
theme_light()
theme_classic()
```



The built in theme functions (`theme_*()`) change the default settings for many elements that can also be changed individually using the`theme()` function. The `theme()` function is a way to further fine-tune the look of your plot. This function takes MANY arguments (just have a look at `?theme`). Luckily there are many great `ggplot` resources online so we don't have to remember all of these, just Google ["ggplot cheat sheet"](https://www.google.com/search?q=ggplot+cheat+sheet&rlz=1C5CHFA_enUS989US989&oq=ggplot+chea&aqs=chrome.0.0i433i512j69i57j0i512l2j0i22i30i395l2.1681j1j7&sourceid=chrome&ie=UTF-8) and find one you like.

Let's look at an example of a `theme()` call, where we change the position of the legend from the right side to the bottom, and remove the ticks of our Locations axis.

```{r}
ggplot(data = daily_visits_loc,
       aes(y = restore_loc, x = daily_visits, fill = visitor_type)) +
    geom_col() +
    labs(x = "Number of Visits",
         y = "Restoration Location",
         fill = "Type of Visitor",
         title = "Total Number of Visits to Delta Restoration Areas by visitor type",
         subtitle = "Sum of all visits during study period") +
    theme_bw() +
    theme(legend.position = "bottom",
          axis.ticks.y = element_blank()) 
    
```


Note that the `theme()` call needs to come after any built-in themes like `theme_bw()` are used. Otherwise, `theme_bw()` will likely override any theme elements that you changed using `theme()`.


You can also save the result of a series of `theme()` function calls to an object to use on multiple plots. This prevents needing to copy paste the same lines over and over again!

```{r save_theme}
my_theme <- theme_bw(base_size = 16) +
    theme(legend.position = "bottom",
          axis.ticks.y = element_blank())
```

So now our code will look like this:


```{r}
ggplot(data = daily_visits_loc,
       aes(y = restore_loc, x = daily_visits, fill = visitor_type)) +
    geom_col() +
    labs(x = "Number of Visits",
         y = "Restoration Location",
         fill = "Type of Visitor",
         title = "Total Number of Visits to Delta Restoration Areas by visitor type",
         subtitle = "Sum of all visits during study period") +
    my_theme

```


::: {.callout-exercise}

What changes do you expect to see in your plot by adding the following line of code? Discuss with your neighbor and then try it out!

`scale_x_continuous(breaks = seq(0,120, 20))`

::: {.callout-answer}
```{r}
ggplot(data = daily_visits_loc,
       aes(y = restore_loc, x = daily_visits, fill = visitor_type)) +
    geom_col() +
    labs(x = "Number of Visits",
         y = "Restoration Location",
         fill = "Type of Visitor",
         title = "Total Number of Visits to Delta Restoration Areas by visitor type",
         subtitle = "Sum of all visits during study period") +
    scale_x_continuous(breaks = seq(0,120, 20)) +
    my_theme
```
:::




Finally we are going to expand the bars all the way to the axis line. In other words, remove the gap between the bars and the vertical "x-axis" line.

```{r}
ggplot(data = daily_visits_loc,
       aes(y = restore_loc, x = daily_visits, fill = visitor_type)) +
    geom_col() +
    labs(x = "Number of Visits",
         y = "Restoration Location",
         fill = "Type of Visitor",
         title = "Total Number of Visits to Delta Restoration Areas by visitor type",
         subtitle = "Sum of all visits during study period") +
    scale_x_continuous(breaks = seq(0,120, 20), expand = c(0,0)) +
    my_theme

```



#### Reordering things

`ggplot()` loves putting things in alphabetical order. But more frequently than not, that's not the order you actually want things to be plotted. One way to do this is to use the `fct_reorder()` function from the `forcats` package. `forcats` provides tools for working with categorical variables. In this case, we want to reorder our categorical variable of `restore_loc` based on the total number of visits.

The first thing we need to do is to add a column to our data with the _total number of visits_ by location. This will be our "sorting" variable.  Then we use `fct_reorder()` to reorder the `restore_loc` variable according to our sorting variable.

```{r}
daily_visits_totals <- daily_visits_loc %>% 
    group_by(restore_loc) %>%
    mutate(n = sum(daily_visits)) %>% 
    ungroup() %>%
    mutate(restore_loc = fct_reorder(restore_loc, n))

head(daily_visits_totals)
levels(daily_visits_totals$restore_loc) ### not alphabetical any more!
```

Next, we will run the code for our plot adding the `fct_reorder()` function.

```{r}
ggplot(data = daily_visits_totals,
       aes(x = daily_visits, y = restore_loc,
           fill = visitor_type)) +
    geom_col() +
    labs(x = "Number of Visits",
         y = "Restoration Location",
         fill = "Type of Visitor",
         title = "Total Number of Visits to Delta Restoration Areas by visitor type",
         subtitle = "Sum of all visits during study period") +
    scale_x_continuous(breaks = seq(0,120, 20), expand = c(0,0)) +
    my_theme

```

What if you want to plot the other way around? In this case from least to greater? We add the `desc()` to the variable we are sorting by.

```{r}
daily_visits_totals <- daily_visits_loc %>% 
    group_by(restore_loc) %>%
    mutate(n = sum(daily_visits)) %>% 
    ungroup() %>%
    mutate(restore_loc = fct_reorder(restore_loc, desc(n)))

ggplot(data = daily_visits_totals,
       aes(x = daily_visits, y = restore_loc,
           fill = visitor_type)) +
    geom_col() +
    labs(x = "Number of Visits",
         y = "Restoration Location",
         fill = "Type of Visitor",
         title = "Total Number of Visits to Delta Restoration Areas by visitor type",
         subtitle = "Sum of all visits during study period") +
    scale_x_continuous(breaks = seq(0,120, 20), expand = c(0,0)) +
    my_theme
```


#### Colors

The last thing we will do to our plot is change the color. To do this we are going to use a function from the `viridis` package. This package provides different color palettes that are designed to improve graph readability for readers with common forms of color blindness and/or color vision deficiency. With `viridis`, there are multiple other color palette packages or color palettes out there that you can use to customize your graphs. We could spend a whole session talking about colors in R! For the purpose of this lesson we are just going to keep it brief and show one function of the `viridis` package that will make our plot colors look better.

```{r}

ggplot(data = daily_visits_totals,
       aes(x = daily_visits, y = restore_loc,
           fill = visitor_type)) +
    geom_col() +
    scale_fill_viridis_d() +
    labs(x = "Number of Visits",
         y = "Restoration Location",
         fill = "Type of Visitor",
         title = "Total Number of Visits to Delta Restoration Areas by visitor type",
         subtitle = "Sum of all visits during study period") +
    scale_x_continuous(breaks = seq(0,120, 20), expand = c(0,0)) +
    my_theme


```

Things to keep in mind when choosing a color palette are the number of variables you have and how many colors your palette has. And if you need a discrete or a continuous color palette. Find more information about colors in this [R color cheatsheet](https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf). 


#### Saving plots

Saving plots using `ggplot` is easy! The `ggsave()` function will save either the last plot you created, or any plot that you have saved to an object. You can specify what output format you want, size, resolution, etc. See `?ggsave()` for documentation.

For example, if we want to save our current plot to an existing folder named "figures", we can do this:

```r
ggsave("figures/visit_restore_site_delta.jpg", width = 12, height = 6, units = "in")
```



#### Creating multiple plots

An easy way to plot another aspect of your data is using the function `facet_wrap()`. This function takes a mapping to a variable using the syntax `~{variable_name}`. The `~` (tilde) is a model operator which tells `facet_wrap()` to model each unique value within `variable_name` to a facet in the plot.

The default behavior of facet wrap is to put all facets on the same x and y scale. You can use the `scales` argument to specify whether to allow different scales between facet plots (e.g `scales = "free_y"` to free the y axis scale). You can also specify the number of columns using the `ncol = ` argument or number of rows using `nrow =`.


```{r}
facet_plot <- ggplot(data = daily_visits_totals,
       aes(x = visitor_type, y = daily_visits,
           fill = visitor_type)) +
    geom_col() +
    facet_wrap(~restore_loc,
               scales = "free_y",
               ncol = 5,
               nrow = 2) +
    scale_fill_viridis_d() +
    labs(x = "Type of visitor",
         y = "Number of Visits",
         title = "Total Number of Visits to Delta Restoration Areas",
         subtitle = "Sum of all visits during study period") +
    theme_bw() +
    theme(legend.position = "bottom",
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank())

facet_plot
```

We can save this plot to our figures folder too. Note that this time we are specifically mentioning the object we want to save.

```r
ggsave("figures/visit_restore_site_facet.jpg", plot = facet_plot, width = 12, height = 8, units = "in")

```


<!--#### Smarter tick labels using `scales`

Fixing tick labels in `ggplot` can be super annoying. The y-axis labels in the plot above don't look great. We could manually fix them, but it would likely be tedious and error prone.

The `scales` package provides some nice helper functions to easily rescale and relabel your plots. Here, we use `scale_y_continuous()` from `ggplot2`, with the argument `labels`, which is assigned to the function name `comma`, from the `scales` package. This will format all of the labels on the y-axis of our plot with comma-formatted numbers.-->


## Interactive visualization

### Tables with `DT`

Now that we know how to make great static visualizations, let's introduce two other packages that allow us to display our data in interactive ways. These packages really shine when used with GitHub Pages, so at the end of this lesson we will publish our figures to the website we created earlier.

First let's show an interactive table of unique sampling locations using `DT`.
We will start by creating a `data.frame` containing unique sampling locations.

```{r uniq_locations}
locations <- visits_long %>%
    distinct(restore_loc, .keep_all = T) %>%
    select(restore_loc, latitude, longitude)

head(locations)
```

The `dplyr::distinct()` function comes pretty handy when you want to filter unique values in a column. In this case we use the `.keep_all = T` argument to keep all the columns of our data frame so we can have the `latitude` and `longitude` of each of the locations. If we don't add this argument, we would end up with a data frame with only one column: `restore_loc` and 10 rows, one for each of the unique locations.


Now we can display this table as an interactive table using `datatable()` from the `DT` package.

```{r display_locations}
datatable(locations)
```

### Maps with `leaflet`

The `leaflet()` package allows you to make basic interactive maps using just a couple lines of code. Note that unlike `ggplot2`, the `leaflet` package uses pipe operators (`%>%`) and not the additive operator (`+`).

The `addTiles()` function without arguments will add base tiles to your map from [OpenStreetMap](https://www.openstreetmap.org/). `addMarkers()` will add a marker at each location specified by the latitude and longitude arguments. Note that the `~` symbol is used here to model the coordinates to the map (similar to `facet_wrap()` in `ggplot`).


```{r simple_map}

leaflet(locations) %>%
    addTiles() %>%
    addMarkers(
        lng = ~ longitude,
        lat = ~ latitude,
        popup = ~ restore_loc
    )
```

<br>

You can also use `leaflet` to import Web Map Service (WMS) tiles. For example, we can use any of the base maps provided by USGS in the [National Map](https://basemap.nationalmap.gov/) archive. For example, let's use the USGSTopo base map. In this example, we also demonstrate how to create a more simple circle marker, the look of which is explicitly set using a series of style-related arguments.


```{r}

leaflet(locations) %>%
    addWMSTiles(
        "https://basemap.nationalmap.gov/arcgis/services/USGSTopo/MapServer/WmsServer",
        layers = "0",
        options = WMSTileOptions(format = "image/png", transparent = TRUE)) %>%
    addCircleMarkers(
        lng = ~ longitude,
        lat = ~ latitude,
        popup = ~ restore_loc,
        radius = 5,
        # set fill properties
        fillColor = "salmon",
        fillOpacity = 1,
        # set stroke properties
        stroke = T,
        weight = 0.5,
        color = "white",
        opacity = 1)

```

<br>

We can also layer base maps. In this case the USGSImageryTopo base map with the USGSHydroCached base map. Note that the url where the map is retrieved is very similar for each USGS base map.
 
```{r}

leaflet(locations) %>%
    addWMSTiles(
        "https://basemap.nationalmap.gov/arcgis/services/USGSImageryTopo/MapServer/WmsServer",
        layers = "0",
        options = WMSTileOptions(format = "image/png", transparent = TRUE)) %>%
    addWMSTiles(
        "https://basemap.nationalmap.gov/arcgis/services/USGSHydroCached/MapServer/WmsServer",
        layers = "0",
        options = WMSTileOptions(format = "image/png", transparent = TRUE)) %>%
    addCircleMarkers(
        lng = ~ longitude,
        lat = ~ latitude,
        popup = ~ restore_loc,
        radius = 5,
        # set fill properties
        fillColor = "salmon",
        fillOpacity = 1,
        # set stroke properties
        stroke = T,
        weight = 0.5,
        color = "white",
        opacity = 1)

```

 

<br>

Leaflet has a ton of functionality that can enable you to create some beautiful, functional maps with relative ease. [Here](https://pages.github.nceas.ucsb.edu/NCEAS/sasap-data/language_vis.html) is an example of some we created as part of the [State of Alaskan Salmon and People (SASAP) project](https://www.nceas.ucsb.edu/workinggroups/state-alaskan-salmon-and-people-sasap), created using the same tools we showed you here. This map hopefully gives you an idea of how powerful the combination of Quarto or RMarkdown and GitHub Pages can be. 


## Publish the Data Visualization lesson to your webpage

::: callout-tip
### Steps

1. Save the `qmd` you have been working on for this lesson.
2. "Render" the `qmd`. This is a good way to test if everything in your code is working.
3. Go to your `index.qmd` and the link to the `html` file with this lesson's content.
4. Save and render `index.qmd` to an `html`.
5. Use the `Git` workflow: ```Stage > Commit > Pull > Push```

:::

## `ggplot2` Resources

- [Why not to use two axes, and what to use instead: The case against dual axis charts](https://blog.datawrapper.de/dualaxis/) by Lisa Charlotte Rost.
- [Customized Data Visualization in `ggplot2`](https://allisonhorst.github.io/eds221-day8-comp/data_viz.html#3_ggplot2_Basics_Review) by Allison Horst.
- [A `ggplot2` tutorial for beautiful plotting in R](https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/#legends) by Cedric Scherer.

