---
title: Data Visualization
---

<!--
Possible prerequisites
- Quarto intro
- R data wrangling
-->

:::{.callout-learning}
After completing this session, you will be able to:

- Describe the four essential components of a `ggplot2::ggplot()` call
- Use `theme` and other functions to customize appealing data visualizations
- Use `facet_wrap()` to create multiple related plots in one figure
- Introduce the `leaflet` and `DT` package to create interactive maps and tables respectively
:::

## Overview

`ggplot2` is a popular package for visualizing data in R.
From the [home page](http://ggplot2.tidyverse.org/):

> `ggplot2` is a system for declaratively creating graphics, based on The Grammar of Graphics. You provide the data, tell `ggplot2` how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details.

The goal of this lesson is to explain the fundamentals of how `ggplot2` work, introduce useful functions for customizing your plots, and inspire you to go and explore this awesome resource for visualizing your data.

::: callout-note

## `ggplot2` vs base graphics in R vs others
There are many different ways to plot your data in R.  All of them can be used to make functional plots! However, `ggplot2` uses a reasonably intuitive syntax to build up a plot in layers and customize it in many different ways. It also has a lot of built in themes and color palettes that can make your plots look nice with just a few lines of code.

Base R graphics  (`plot()`, `hist()`, etc) can be helpful for simple, quick-and-dirty plots, e.g., for data exploration, but it can be hard to decipher and customize. `ggplot2` can be used for almost everything else, including publication-ready graphics.

:::

## Setup

Make sure you’re in the right project (`training_<username>`). Use Git Pull to check for any changes. Then, create a new Quarto document, delete the default text below the YAML header, and save this document.  

:::{.callout-tip collapse=true}
## Quarto header options?

Consider adding Quarto options to the YAML header to control the output:
```yaml
format:
  html:
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  warning: false
  message: false
```
:::

### Load Packages

Load the packages we'll need. They're grouped by purpose to make it easier to remember what each one does.

```{r load_packages}
#| message: false
#| warning: false

### Data wrangling
library(readr)
library(dplyr)
library(tidyr)
library(forcats)   ### tools for working with factors (e.g., reordering)
library(janitor)   ### clean column names

### Visualization
library(ggplot2)
library(scales)    ### axis formatting helpers
library(viridis)   ### colorblind-friendly palettes

### Interactive output
library(leaflet)
library(DT)
```

### Load the Data

Download the data from the [EDI Data Repository](https://portal.edirepository.org/nis/mapbrowse?packageid=edi.587.1): scroll to *Resources*, click the "Socioecological monitoring data" download button, and save to your `data/` folder. Then read it in:

```{r}
#| echo: false
#| eval: true

### NOTE THIS CHUNK IS HIDDEN (echo: false) AND EVALUATED (eval: true)
### FOR THE PURPOSES OF RENDERING THE LESSON, WITH THE EXTRA FOLDER IN PATH.
### THE FOLLOWING CHUNK (echo: true and eval: false) IS FOR WHERE USERS SHOULD
### SAVE FILES.

delta_visits_raw <- read_csv(file.path("data", "r_data_visualization", 
                                       "Socioecological_monitoring_data.csv"),
                             show_col_types = FALSE)
```

```{r}
#| echo: true
#| eval: false

# read in data from the data directory after manually downloading data 
delta_visits_raw <- read_csv("data/Socioecological_monitoring_data.csv")

```

### Learn About the Data

For this session we are going to be working with data on **Socioecological Monitoring on the Sacramento-San Joaquin Delta**. Check out [the documentation](https://portal.edirepository.org/nis/mapbrowse?packageid=edi.587.1): read the abstract, explore the metadata, check the intellectual rights, etc. 

### Explore the Data

Finally, let's explore the data we just read into our working environment. 

```r
### Check out column names
colnames(delta_visits_raw)

### Peek at each column and class
glimpse(delta_visits_raw)

### From when to when
range(delta_visits_raw$Date)

### Which time of day?
unique(delta_visits_raw$Time_of_Day)
```


## Prepare the Data

We nearly always need to do some wrangling before we can analyze and plot our data.  Here we will wrangle our data into a tidy format that is `ggplot2`-friendly.

### Clean Column Names

`janitor::clean_names()` converts all column names to consistent `snake_case` in a single line:

```{r}
colnames(delta_visits_raw)[1:6]

delta_visits <- delta_visits_raw %>%
  janitor::clean_names()

colnames(delta_visits)[1:6]
```

### Reshape to Long (Tidy) Format

Recall the [tidy data principles](https://r4ds.hadley.nz/data-tidy.html#fig-tidy-structure){.external target="_blank"}.

1. Each variable is a column; each column is a variable.
2. Each observation is a row; each row is an observation.
3. Each value is a cell; each cell is a single value.

Which (if any) of these principles are violated in our data frame `delta_visits`?

`ggplot2` works best with **tidy data** — one row per observation. Our visitor-type counts are spread across multiple columns, so we pivot them (using `tidyr::pivot_longer()`) into a single `visitor_type` column:

```{r}
visits_long <- delta_visits %>%
  tidyr::pivot_longer(
    cols = c(sm_boat, med_boat, lrg_boat, bank_angler, scientist, cars),
    names_to = "visitor_type",
    values_to = "quantity"
  ) %>%
  ### shorten the long name:
  dplyr::rename(restore_loc = eco_restore_approximate_location) %>%
  ### drop the notes column:
  dplyr::select(-notes)

head(visits_long)
```



::: callout-exercise

Using the `visits_long` dataframe, calculate the daily visits by `restore_loc`, `date`, and `visitor_type`.  Assign the result to a new object called `daily_visits_loc`.  Then inspect the result.

::::{.callout-answer}
 
```{r mean_esc}
daily_visits_loc <- visits_long %>%
    dplyr::group_by(restore_loc, date, visitor_type) %>% 
    dplyr::summarize(daily_visits = sum(quantity), .groups = 'drop')
    
head(daily_visits_loc)
```

Refresher:

* `group_by()` to calculate our results for the unique combinations of type of visit, restoration location proximity, and day.  
* `summarise()` to sum up the daily visit value for each of these groups. 
* `%>%` operator to pipe in the result of one command as an argument to the next one.

::::

:::

## Plot With `ggplot2`

### Essential Components

Every `ggplot2` call requires four essential components:

1. **`ggplot()`** opens the plot (note, the **package** is called `ggplot2` but the function is just `ggplot()`.)
2. **`data =`** defines the data frame to use
3. **`aes()`** maps variables to visual properties (axes, color, fill, etc.)
4. **`geom_*()`** defines the plot type (bar, point, line, box, etc.)

Layers are added with `+`.

```{r}
#| fig-align: "center"
#| fig-height: 3.5

ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits)) +
  geom_col()
```

::: callout-tip
`data =` and `aes()` can also live inside the `geom_*()` call — useful when layering multiple data sources!  But for simple plots, a good practice is to keep them in `ggplot()` for clarity.

**Variations on `data=` and `aes()`**:  Options 1 - 3 produce the same output as above.  Option 4 flips the axes.  Examine the code and the outputs.

:::: panel-tabset
## Option 1 (original)

Data and mapping defined in `ggplot()` function:
```{r}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5

ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits)) +
    geom_col()
```

## Option 2

Data defined in `ggplot()` function; mapping defined in `geom_*()`:
```{r}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5
ggplot(data = daily_visits_loc) +
    geom_col(aes(x = restore_loc, 
                 y = daily_visits))
```

## Option 3

Data and mapping both defined in `geom_*()`
```{r}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5
ggplot() +
    geom_col(data = daily_visits_loc,
             aes(x = restore_loc, y = daily_visits))
```

## Option 4 (axes flipped)

Data and mapping both defined in `geom_*()`, but with x and y flipped.  Why might you prefer this?
```{r}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5
ggplot() +
    geom_col(data = daily_visits_loc,
             aes(y = restore_loc, x = daily_visits))
```

:::: 
<!-- end tabset -->
:::
<!-- end tip -->


### Different `geom_*` Types

Having the basic structure with the essential components in mind, we can easily change the type of graph by updating the `geom_*()`.  First we will filter the data to remove some extreme values, and to focus on boat users specifically.

```{r}
daily_boats_filtered <- daily_visits_loc %>%
    filter(daily_visits < 30,
           visitor_type %in% c("sm_boat", "med_boat", "lrg_boat"))
```

:::panel-tabset

## Boxplot

```{r boxplot_ex}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5
ggplot(data = daily_boats_filtered, aes(x = visitor_type, y = daily_visits)) +
    geom_boxplot()
```


## Violin Plot

Note, axes flipped to make it easier to read the labels.
```{r violin_plot}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5
ggplot(data = daily_boats_filtered, aes(x = daily_visits, y = visitor_type)) +
    geom_violin()
```



## Boxplot With Jitter

Axes flipped, and all observations now plotted using `geom_jitter()`, which adds a small amount of random variation to the location of each point, to prevent overplotting.  Note the `geom_*()` functions have additional parameters to control appearance.
```{r boxplot_w_jitter}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5
ggplot(data = daily_boats_filtered, aes(x = daily_visits, y = visitor_type)) +
    geom_boxplot() +
    geom_jitter(height = .1)
```


## Scatterplot

Note, new aesthetics: `color` is now mapped to `visitor_type`.  Note the legend!
```{r line_poin_plot}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5
ggplot(data = daily_boats_filtered, aes(x = date, y = daily_visits)) +
    geom_point(aes(color = visitor_type))
```
:::
<!-- end tabset -->

## Customize the Plot

We'll build our final bar chart in two focused steps: **labels & theme**, then **color & ordering**.

### Set Labels, Axes, and Themes

::: {.callout-tip}
## Inside vs. outside `aes()`

- **Inside `aes()`**: maps a *variable* to an aesthetic:  color tracks the data.  
- **Outside `aes()`**: sets a *constant*: all elements get the same color.

::::panel-tabset

## Right: Set Fixed Color

Set `fill` to a specific, valid color outside `aes()`.  All bars are the same color!
```{r}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5
ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits)) +
  geom_col(fill = "steelblue")
```

## Right: Map Color to Variable

Set `fill` to map to a single variable, inside `aes()` (whether the result makes sense, that's on you)
```{r}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5
ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits)) +
  geom_col(aes(fill = visitor_type))
```

## Wrong! `aes(fill="blue")`

Setting to a specific color **inside** the `aes()` function is incorrect.  This doesn't throw an error, but clearly does not reflect what you might be hoping for!
```{r}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5
ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits, fill = "blue")) +
  geom_col()
```

::::
<!-- end tabset -->

:::
<!-- end callout -->


To our basic bar chart, let's clean it up using some of the above ideas, and some new ones:

* Map `visitor_type` to `fill`
* flip the axes so location labels are readable
* add informative labels
* apply a clean theme

```{r}
#| out-width: "80%"
#| fig-align: "center"
#| fig-height: 3.5
ggplot(data = daily_visits_loc,
       aes(y = restore_loc, x = daily_visits, fill = visitor_type)) +
  geom_col() +
  labs(x     = "Number of Visits",                                        # <1>
       y     = "Restoration Location",                                    # <1>
       fill  = "Visitor Type",                                            # <1>
       title = "Total Visits to Delta Restoration Areas by Visitor Type", # <1>
       subtitle = "Sum of all visits July 2017 - March 2018") +           # <1>
  scale_x_continuous(breaks = seq(0, 120, 20), expand = c(0, 0)) +        # <2>
  theme_minimal() +                                                            # <3>
  theme(
    legend.position = "bottom",                                           # <4>
    axis.ticks.y    = element_blank()                                     # <5>
  )
```
1. `labs()` to rename axes, legend, title, subtitle, etc.
2. `scale_x_continuous()` to set x-axis breaks & labels, and remove padding (expand)
3. `theme_minimal()`, `theme_bw()`, and other built-in themes for a clean, publication-ready plot
4. `theme()` to customize specific theme elements (e.g., legend position)
5. `element_blank()` to effectively turn off any theme element

:::callout-tip

## Built-in and custom themes

Always put `theme()` *after* built-in themes like `theme_minimal()`, or the built-in theme will override your tweaks. 

You can also create your own custom theme object to reuse it across plots, just as you would a built-in theme - just wrap it in a function!
```r
my_theme <- function(size = 14) {
    theme_minimal(base_size = size) +
    theme(legend.position = "bottom", 
          axis.ticks.y = element_line(color = "red", size = 0.5))
}
```
:::

### Order categorical values and add color

For categorical variables (e.g., text), `ggplot2` defaults to displaying values in alphabetical order (see the plots above). We can use factors (ordered vectors of categorical values) to force a particular display order.  One easy approach is to use `forcats::fct_reorder()` from `forcats` to sort by a meaningful variable instead.

```{r}
daily_visits_totals <- daily_visits_loc %>%
  group_by(restore_loc) %>%
  mutate(total = sum(daily_visits)) %>%
  ungroup() %>%
  mutate(restore_loc = fct_reorder(restore_loc, desc(total)))

head(daily_visits_totals$restore_loc)
```

Note that `fct_reorder()` defaults to sorting in ascending order, so the location with the lowest total visits will be at the bottom of the plot.  To reverse this, add `desc()` around the variable you are sorting by: `fct_reorder(restore_loc, desc(total))`.  Note how `Prospect` is the first level of the factor, and has the highest total visits.

Default colors in `ggplot2` are not the most appealing, and not ideal for accessibility.  There are many options available, and colors in ggplot could be an entire topic in itself.  Here are some quick tips:

::: panel-tabset
## Color scales

Use `scale_color_*()` or `scale_fill_*()` to select or modify color palettes.  Good choices:

* `scale_color_viridis_c/d()`: {{< fa thumbs-up >}} {{< fa thumbs-up >}} Colorblind-friendly, perceptually uniform (`c`ontinuous or `d`iscrete)
* `scale_color_brewer()`: [ColorBrewer](https://colorbrewer2.org/){.external target="_blank"} palettes (categorical, sequential, diverging)
* `scale_color_gradient()`: Custom two-color gradient
* `scale_color_gradient2()`: Diverging gradient with midpoint
* `scale_color_manual()`: Fully custom discrete colors

## Color name vs code

You can use either color names (e.g., "steelblue", "salmon", "orchid") or hex codes (e.g., "#4682B4", "#FA8072", "#DA70D6") to specify colors in R.  Hex codes are more precise and allow for a wider range of colors, but color names can be easier to remember and read in code. 

Use `colors()` to see all available color names in R, or [check out this cheat sheet](https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf){.external target="_blank"}.

## Alpha

Use the `alpha` argument to set transparency, which can help with overplotting.  For example, `geom_point(alpha = 0.5)` will make points semi-transparent.

:::


Let's plot our daily visits data, with the locations ordered by total visits (from `fct_reorder()` above), and colored by visitor type.  We'll add a colorblind-friendly palette with `scale_fill_viridis_d()`.

```{r}
my_theme <- function(base_size = 14) {
    theme_minimal(base_size = base_size) +
    theme(legend.position = "bottom", 
          panel.grid.major = element_line(color = 'slateblue1'))
}

ggplot(data = daily_visits_totals,
       aes(y = restore_loc, x = daily_visits, fill = visitor_type)) +
  geom_col() +
  scale_fill_viridis_d() +
  scale_x_continuous(breaks = seq(0, 120, 20), expand = c(0, 0)) +
  labs(
    x     = "Number of Visits",
    y     = "Restoration Location",
    fill  = "Visitor Type",
    title = "Total Visits to Delta Restoration Areas by Visitor Type",
    subtitle = "Sum of all visits during the study period"
  ) +
  my_theme()
```

:::callout-exercise
Why is the longest bar at the bottom? It might make more sense to put the longest bar on top - how could we do that?

::::{.callout-answer}
The y axis starts with the lowest value on the bottom, and here the "lowest" value is the first factor level (which corresponds to the *highest* total visits as we defined it in an earlier code block). To put the longest bar on top, we can reverse the order of the factor levels in `restore_loc` using `forcats::fct_reorder()` and removing the `desc()` wrapper we included earlier.
::::
:::

### Save your plot

```{r}
#| eval: false
ggsave("figures/visit_restore_site_delta.jpg", width = 12, height = 6, units = "in")
```

`ggsave()` automatically saves the last plot displayed, or pass `plot = my_object` to save a specific one.  Set `width`, `height`, and `units` to control the size of the output file.  By default, `ggsave()` guesses the output format from the file extension (e.g., `.jpg`, `.png`, `.pdf`).  See `?ggsave()` for more options.



## Creating multiple plots

An easy way to plot another aspect of your data is using the function `facet_wrap()`. This function takes a mapping to a variable using the syntax `~{variable_name}`. The `~` (tilde) is a model operator which tells `facet_wrap()` to model each unique value within `variable_name` to a facet in the plot.

The default behavior of facet wrap is to put all facets on the same x and y scale. You can use the `scales` argument to specify whether to allow different scales between facet plots (e.g `scales = "free_y"` to free the y axis scale). You can also specify the number of columns using the `ncol = ` argument or number of rows using `nrow =`.


```{r}
facet_plot <- ggplot(data = daily_visits_totals,
       aes(x = visitor_type, y = daily_visits,
           fill = visitor_type)) +
    geom_col() +
    facet_wrap(~restore_loc,
               scales = "free_y",
               ncol = 5,
               nrow = 2) +
    scale_fill_viridis_d() +
    labs(x = "Type of visitor",
         y = "Number of Visits",
         title = "Total Number of Visits to Delta Restoration Areas",
         subtitle = "Sum of all visits during study period") +
    theme_bw() +
    theme(legend.position = "bottom",
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank())

facet_plot
```

We can save this plot to our figures folder too. Note that this time we are specifically mentioning the object we want to save.

```r
ggsave("figures/visit_restore_site_facet.jpg", plot = facet_plot, width = 12, height = 8, units = "in")

```


<!--#### Smarter tick labels using `scales`

Fixing tick labels in `ggplot` can be super annoying. The y-axis labels in the plot above don't look great. We could manually fix them, but it would likely be tedious and error prone.

The `scales` package provides some nice helper functions to easily rescale and relabel your plots. Here, we use `scale_y_continuous()` from `ggplot2`, with the argument `labels`, which is assigned to the function name `comma`, from the `scales` package. This will format all of the labels on the y-axis of our plot with comma-formatted numbers.-->


## Interactive visualization

### Tables with `DT`

Now that we know how to make great static visualizations, let's introduce two other packages that allow us to display our data in interactive ways. These packages really shine when used with GitHub Pages, so at the end of this lesson we will publish our figures to the website we created earlier.

First let's show an interactive table of unique sampling locations using `DT`.
We will start by creating a `data.frame` containing unique sampling locations.

```{r uniq_locations}
locations <- visits_long %>%
    distinct(restore_loc, .keep_all = T) %>%
    select(restore_loc, latitude, longitude)

head(locations)
```

The `dplyr::distinct()` function comes pretty handy when you want to filter unique values in a column. In this case we use the `.keep_all = T` argument to keep all the columns of our data frame so we can have the `latitude` and `longitude` of each of the locations. If we don't add this argument, we would end up with a data frame with only one column: `restore_loc` and 10 rows, one for each of the unique locations.


Now we can display this table as an interactive table using `datatable()` from the `DT` package.

```{r display_locations}
datatable(locations)
```

### Maps with `leaflet`

The `leaflet()` package allows you to make basic interactive maps using just a couple lines of code. Note that unlike `ggplot2`, the `leaflet` package uses pipe operators (`%>%`) and not the additive operator (`+`).

The `addTiles()` function without arguments will add base tiles to your map from [OpenStreetMap](https://www.openstreetmap.org/). `addMarkers()` will add a marker at each location specified by the latitude and longitude arguments. Note that the `~` symbol is used here to model the coordinates to the map (similar to `facet_wrap()` in `ggplot`).


```{r simple_map}

leaflet(locations) %>%
    addTiles() %>%
    addMarkers(
        lng = ~ longitude,
        lat = ~ latitude,
        popup = ~ restore_loc
    )
```

<br>

You can also use `leaflet` to import Web Map Service (WMS) tiles. For example, we can use any of the base maps provided by USGS in the [National Map](https://basemap.nationalmap.gov/) archive. For example, let's use the USGSTopo base map. In this example, we also demonstrate how to create a more simple circle marker, the look of which is explicitly set using a series of style-related arguments.


```{r}

leaflet(locations) %>%
    addWMSTiles(
        "https://basemap.nationalmap.gov/arcgis/services/USGSTopo/MapServer/WmsServer",
        layers = "0",
        options = WMSTileOptions(format = "image/png", transparent = TRUE)) %>%
    addCircleMarkers(
        lng = ~ longitude,
        lat = ~ latitude,
        popup = ~ restore_loc,
        radius = 5,
        # set fill properties
        fillColor = "salmon",
        fillOpacity = 1,
        # set stroke properties
        stroke = T,
        weight = 0.5,
        color = "white",
        opacity = 1)

```

<br>

We can also layer base maps. In this case the USGSImageryTopo base map with the USGSHydroCached base map. Note that the url where the map is retrieved is very similar for each USGS base map.
 
```{r}

leaflet(locations) %>%
    addWMSTiles(
        "https://basemap.nationalmap.gov/arcgis/services/USGSImageryTopo/MapServer/WmsServer",
        layers = "0",
        options = WMSTileOptions(format = "image/png", transparent = TRUE)) %>%
    addWMSTiles(
        "https://basemap.nationalmap.gov/arcgis/services/USGSHydroCached/MapServer/WmsServer",
        layers = "0",
        options = WMSTileOptions(format = "image/png", transparent = TRUE)) %>%
    addCircleMarkers(
        lng = ~ longitude,
        lat = ~ latitude,
        popup = ~ restore_loc,
        radius = 5,
        # set fill properties
        fillColor = "salmon",
        fillOpacity = 1,
        # set stroke properties
        stroke = T,
        weight = 0.5,
        color = "white",
        opacity = 1)

```

 

<br>

Leaflet has a ton of functionality that can enable you to create some beautiful, functional maps with relative ease. [Here](https://pages.github.nceas.ucsb.edu/NCEAS/sasap-data/language_vis.html) is an example of some we created as part of the [State of Alaskan Salmon and People (SASAP) project](https://www.nceas.ucsb.edu/workinggroups/state-alaskan-salmon-and-people-sasap), created using the same tools we showed you here. This map hopefully gives you an idea of how powerful the combination of Quarto or RMarkdown and GitHub Pages can be. 


## Publish the Data Visualization lesson to your webpage

::: callout-tip
### Steps

1. Save the `qmd` you have been working on for this lesson.
2. "Render" the `qmd`. This is a good way to test if everything in your code is working.
3. Go to your `index.qmd` and the link to the `html` file with this lesson's content.
4. Save and render `index.qmd` to an `html`.
5. Use the `Git` workflow: ```Stage > Commit > Pull > Push```

:::

## `ggplot2` Resources

- [Why not to use two axes, and what to use instead: The case against dual axis charts](https://blog.datawrapper.de/dualaxis/) by Lisa Charlotte Rost.
- [Customized Data Visualization in `ggplot2`](https://allisonhorst.github.io/eds221-day8-comp/data_viz.html#3_ggplot2_Basics_Review) by Allison Horst.
- [A `ggplot2` tutorial for beautiful plotting in R](https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/#legends) by Cedric Scherer.

