---
title: Writing Functions
---

```{r}
#| label: load-libs-funcs
#| message: false
#| warning: false
#| echo: false

library(DT)
```

:::{.callout-learning}
After completing this session, you will be able to:

-   Explain the importance of using and developing functions
-   Create custom functions using R code
-   Document functions to improve understanding and code communication
:::

## Why Functions?

Many R users write code as a continuous stream of commands—paste from the console, run, repeat. That works, but breaking code into small, reusable **functions** pays off quickly.

The guiding principle is **DRY: Don't Repeat Yourself**. If you've copied and pasted a block of code more than twice, it's time to write a function. Functions give you:

- **Less repetition** — change logic in one piece of code, not ten
- **Fewer errors** — one tested implementation instead of many copies
- **Better readability** — a well-named function documents the coder's intent

### Function Basics

The syntax for defining a function in R is:

```r
my_function <- function(argument1, argument2) {
  <code that performs some well defined task>
  return(result)
}
```

You can call it just like any built-in function:

```r
result <- my_function(argument1 = value1, argument2 = value2)
```

::: {.callout-tip}
## Naming Functions
Ideally, function names should be short, but still clearly capture what the function does.

Best Practices from [Chapter 19 Functions](https://r4ds.had.co.nz/functions.html) in R for Data Science:

- **Function names** should be **verbs** and **arguments** should be **nouns** (there are [exceptions](https://r4ds.had.co.nz/functions.html#functions-are-for-humans-and-computers)).
- Use the `snake_case` naming convention for functions that are multiple words.
- For a "family" of functions, use a common prefix to indicate that they are connected. For example, most of the functions in the `usethis` package have the prefix `use_*`

Based on these tips, is `my_function()` a good name for a function?
:::

### Example: Temperature Conversion 

Imagine you have some temperature data measured in degrees Fahrenheit and you want to convert that to Celsius for your analysis. You might have an R script that does this for you.

```{r}
#| label: temp-test-data

airtemps <- c(212, 30.3, 78, 32)
celsius1 <- (airtemps[1] - 32) * 5/9
celsius2 <- (airtemps[2] - 32) * 5/9
celsius3 <- (airtemps[3] - 33) * 5/9
```

Here we repeat the same formula three times.  And note the third time, there's a subtle mistake!  This code would be both more compact and more reliable if we didn't repeat ourselves.

#### Convert Fahrenheit to Celsius {.unnumbered}

To create a function in R, we use the `function` function (so meta!) and assign its result to an object. Let's create a function that calculates Celsius temperature outputs from Fahrenheit temperature inputs.

```{r}
#| label: f2c-function

convert_f_to_c <- function(fahr) {
  celsius <- (fahr - 32) * 5/9
  return(celsius)
}
```

Because R operations are vectorized, this works on a single value *or* an entire vector:

```{r}
### check single value:
celsius1a <- convert_f_to_c(airtemps[1])
celsius1a == celsius1
### calculate for whole vector:
celsius_vec <- convert_f_to_c(airtemps)
celsius_vec
```


::: {.callout-exercise}
## Convert Celsius to Fahrenheit

Write a function `convert_c_to_f` that reverses the conversion. Use it to convert `celsius` back to Fahrenheit and verify that the result matches `airtemps`.

**Hint:** the formula for Celsius to Fahrenheit conversions is `celsius * 9/5 + 32`.

:::: {.callout-answer}

```{r}
#| label: f2c-func-solution

convert_c_to_f <- function(celsius) {
    fahr <- celsius * 9/5 + 32
    return(fahr)
}

fahr_vec <- convert_c_to_f(celsius_vec)
airtemps == fahr_vec
```

::::
<!-- end answer -->

:::
<!-- end exercise -->

## Function Arguments and `return`

### Argument Defaults

Function arguments often include a default value - in which case, the user can opt to simply not assign a value.  Here, `convert_temp()` allows the user to specify the scale of their input, defaulting to Fahrenheit (`scale = 'f'`).

:::callout-tip
## Error handling!

Sometimes a user will include an argument that breaks the code - a character when the function expects a number, upper-case when expecting lower-case, etc.  Try to anticipate simple errors and include a way to identify and handle them in the function: try to correct the error on the fly, or use `stop()` to return a useful error message to the user.
:::

```{r}
convert_temp <- function(t, scale = 'f') {
    scale = tolower(substr(scale, 1, 1))         # <1>
    if(!scale %in% c('c', 'f')) {                # <2>
      stop('scale must be either "c" or "f"')
    }
    if(scale == "c") {                           # <3>
        result = t * 9/5 + 32
    } else {
        result  = (t - 32) * 5/9
    }
    return(round(result, 3))
}
```
1. What if user enters `scale = "Celsius"`?  `substr(x, 1, 1)` grabs just the first letter, and `tolower` forces it to lower case
2. If `scale` is not one of the expected values, generate a sensible error with `stop()`
3. We know the values are valid now, so we can use a simple if/else to do the correct calculation!

```{r}
convert_temp(t = airtemps)
convert_temp(t = airtemps, scale = 'Celsius')
```

### Return Values

Use `return(x)` to end the function and hand the result back to the user.  To provide multiple values back to a user, consider more complex data structures like a dataframe (well structured) or a named list (very flexible).

::: {.panel-tabset}

## Dataframe

```{r}
convert_f_to_c_k <- function(f) {
  c <- (f - 32) * 5/9
  k  <- c + 273.15
  out_df <- data.frame(fahr = f, celsius = c, kelvin = k)
  return(out_df)
}

convert_f_to_c_k(seq(-100, 100, 50))
```

## Named List

```{r}
convert_f_to_c_k <- function(f) {
  c <- (f - 32) * 5/9
  k  <- c + 273.15
  out_list <- list(fahr = f, celsius = c, kelvin = k)
  return(out_list)
}

convert_f_to_c_k(seq(-100, 100, 50))
```

:::

:::callout-tip
## Implicit vs explicit `return()`

If you don't explicitly state a value to `return()`, R will pass the result of the last step of the function.  For simple functions, this is fine, but for anything more than a line or two, best practice is to explicitly call `return()` so it is obvious what is being returned.
:::



## Documenting Functions

Well-named functions are a start, but good documentation tells collaborators (and future you) what a function expects and what it returns.  Comments in the function body are a good start.  For a more standardized structure, the [`roxygen2`](https://roxygen2.r-lib.org/) package provides a lightweight format for this. Place comments starting with `#'` immediately above the function definition, and use specific tags (e.g., `@param`) to define sections of the documentation.

```r
#' Convert temperature from Fahrenheit to Celsius
#'
#' @param fahr Numeric value or vector in degrees Fahrenheit
#'
#' @return Numeric value or vector in degrees Celsius
#' @export
#'
#' @examples
#' convert_f_to_c(32)
#' convert_f_to_c(c(32, 212, 72))
convert_f_to_c <- function(fahr) {
  celsius <- (fahr - 32) * 5/9
  return(celsius)
}
```

Key tags:

| Tag | Purpose |
|---|---|
| `@param` | Describes each input argument to the function |
| `@return` | Describes the function's output |
| `@examples` | Shows usage examples |
| `@export` | Makes the function available if bundled in a package |

This `roxygen2` structure might seem overly complicated, but it will be useful for others interested in using your functions (including "future you") - especially when including your functions in an R package!

::: {.callout-note}
For more best practices on function documentation, check out Hadley Wickham and Jennifer Bryan's online book [R Packages (2e)](https://r-pkgs.org/) - [Chapter 10, Section 16: Function Documentation](https://r-pkgs.org/man.html).
:::


### Exercise: Minimizing Work with Functions

Functions can of course be as simple or complex as needed. They can be be very effective in repeatedly performing calculations, or for bundling a group of commands that are used on many different input data sources. For example, we might create a simple function that takes Fahrenheit temperatures as input, and calculates both Celsius and Kelvin temperatures. All three values are then returned in a list, making it very easy to create a comparison table among the three scales.

```{r}
#| label: func-return-list

convert_temps <- function(fahr) {
  celsius <- (fahr - 32) * 5/9
  kelvin <- celsius + 273.15
  return(list(fahr = fahr, celsius = celsius, kelvin = kelvin))
}

temps_df <- data.frame(convert_temps(seq(-100,100,10)))
```

```{r}
#| label: list-return-table
#| echo: false

datatable(temps_df)
```

Once we have a dataset like that, we might want to plot it. One thing that we do repeatedly is set a consistent set of display elements for creating graphs and plots. By using a function to create a custom `ggplot` theme, we can enable to keep key parts of the formatting flexible. For example, in the `custom_theme` function, we provide a `base_size` argument that defaults to using a font size of 9 points. Because it has a default set, it can safely be omitted. But if it is provided, then that value is used to set the base font size for the plot.

```{r}
#| label: func-custom-theme

custom_theme <- function(base_size = 9) {
    ggplot2::theme(
      text             = ggplot2::element_text(family = 'Helvetica', 
                                               color = 'gray30', 
                                               size = base_size),
      plot.title       = ggplot2::element_text(size = ggplot2::rel(1.25), 
                                               hjust = 0.5, 
                                               face = 'bold'),
      panel.background = ggplot2::element_blank(),
      panel.border     = ggplot2::element_blank(),
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major = ggplot2::element_line(colour = 'grey90', 
                                               linewidth = 0.25),
      legend.position  = 'right',
      legend.key       = ggplot2::element_rect(colour = NA, 
                                               fill = NA),
      axis.ticks       = ggplot2::element_blank(),
      axis.line        = ggplot2::element_blank()
      )
}

library(ggplot2)

ggplot(temps_df, mapping = aes(x = fahr, y = celsius, color = kelvin)) +
    geom_point() +
    custom_theme(10)

```

In this case, we set the font size to 10, and plotted the air temperatures. The `custom_theme` function can be used anywhere that one needs to consistently format a plot.

But we can go further. One can wrap the entire call to ggplot in a function, enabling one to create many plots of the same type with a consistent structure. For example, we can create a `scatterplot` function that takes a data frame as input, along with a point_size for the points on the plot, and a font_size for the text.

```{r}
#| label: func-plot-custom

scatterplot <- function(df, point_size = 2, font_size = 9) {
  ggplot(df, mapping = aes(x = fahr, y = celsius, color = kelvin)) +
    geom_point(size = point_size) +
    custom_theme(font_size)
}
```

Calling that let's us, in a single line of code, create a highly customized plot but maintain flexibility via the arguments passed in to the function. Let's set the point size to 3 and font to 16 to make the plot more legible.

```{r}
#| label: func-call-sp

scatterplot(temps_df, point_size = 3, font_size = 16)
```

Once these functions are set up, all of the plots built with them can be reformatted by changing the settings in just the functions, whether they were used to create 1, 10, or 100 plots.

:::{.callout-tip}
### Summary

-   Functions are useful to reduce redundancy, reuse code, and reduce errors
-   Build functions with `function()`
-   Document functions with `roxygen2` comments

:::

::: {.callout-tip}
#### Workflow for Creating Functions

1. Have a clear goal (sometimes it helps to create a visual).
2. Outline the plan and then add more detailed steps or tasks.
3. Build it up bit-by-bit and start with a minimum viable example. As your function becomes more complex, it can harder to track all the bits.
4. Always check intermediates!
:::
