---
title: Creating R Packages
---

:::{.callout-learning}
After completing this session, you will be able to:

- Discuss the advantages of creating your own R package to organize code
- Identify simple techniques for creating an R package
- Create CRAN-compliant documentation for functions in a package
:::

## Overview

Most R users are familiar with loading and using packages in their work; we've worked with a number of well-known packages in this course thus far! In addition, even if they mostly stick to a core set of packages, they understand the breadth of [packages available on CRAN](https://cran.r-project.org/web/packages/available_packages_by_name.html) (the R package authority) for handling virtually any conceivable need. However, most R users have never created a package for their own work, often out of a fear that the process is too complicated. In reality, it's fairly straighforward and super useful in your personal work. Creating packages serves two main use-cases:

- Distribute reusable code among projects (even if just for yourself)
- Reproducibly document analysis and models and their results

Even if you don't plan on writing a package with such broad appeal such as, say, `ggplot2` or `dplyr`, you still might consider creating a package to contain:

- Useful utility functions you write (i.e., a [Personal Package](https://hilaryparker.com/2013/04/03/personal-r-packages/))
  - Having a place to put these functions makes it much easier to find and use them later.
- A set of shared routines for your lab or research group, making it easier to remain consistent within your team and also to save time.
- The analysis accompanying a thesis or manuscript, making it all that much easier for others to reproduce your results.

## Get Ready

Before we get started with the tutorial, go ahead and **make a new folder/project for the R package we'll start today.** Name that project/folder like so: `pkg_{lastname}` (e.g., `pkg_ohara` or `pkg_lyon`). 

:::{.callout-warning icon="false"}
### Packages & GitHub
Make sure that your package is also tracked by Git and linked to a remote repository of the same name in GitHub! Include a ".gitignore" as you normally would for any other such project.

Users--including you--can install your package from GitHub even if you don't go all the way to getting it accepted by CRAN.
:::

### Install Packages for Maintaining Packages

The `usethis`, `devtools`, `roxygen2`, and `testthat` packages greatly streamline the process of creating and maintaining a package. You should install all of these now if you have not already done so.

```{r}
#| label: pkgs-for-pkgs
#| eval: false

# install.packages("librarian")
librarian::shelf(usethis, devtools, roxygen2, testthat)
```

## Create Core Package Skeleton

Fundamentally, an R package is a bundle of scripts that can be installed on any computer in order to grant access to those scripts. The core package architecture is intended to facilitate the installation step (so that `install.packages` works as expected, for example) and lets people who install the package use the functions as they would functions from any other package.

**For an R package to work (i.e., be install-able), you need a few core components.** Click through the tabs below--in order--to create the fundamental structure for your R package.

:::{.panel-tabset}
### 1. "DESCRIPTION"

Make a new file named _exactly_ "DESCRIPTION" (all caps, no file extension). This file is the central pillar of any R package as it tells R that this folder is a package along with a number of other necessary pieces of information (e.g., what other packages need to be installed, what R version is required, the license).

Once you have this file, **click the collapsed menu below and copy/paste all of its contents into your empty DESCRIPTION file.**

<details> 
<summary>DESCRIPTION Contents</summary>

Change the part of the first line after "Package: " to match the name of your folder.

```{r}
#| label: make-description
#| eval: false

Package: pkg_{lastname}
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R: 
    person("First", "Last", , "first.last@example.com", role = c("aut", "cre"))
Description: What the package does (one paragraph).
License: `use_mit_license()`, `use_gpl3_license()` or friends to pick a
    license
Encoding: UTF-8
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.3.3
```

</details>

### 2. ".Rbuildignore"

Make a new file named _exactly_ ".Rbuildignore", ignore any warnings about creating a file that starts with a period. This file is to an R package as the .gitignore is to a Git repository; it tells the package what files in your folder should not be included in the R package. This is typically used to avoid including things that would break your package or if you have content necessary for the package build on your (i.e., the maintainer's) end, but not useful to the end user.

Once you have this file, **click the collapsed menu below and copy/paste all of its contents into your empty .Rbuildignore.**

<details> 
<summary>.Rbuildignore Contents</summary>

Technically you only _need_ these two files listed if you're doing this in RStudio but it doesn't hurt to include them at this point and is a good excuse to make a necessary file.
```{r}
#| label: make-rbuildignore
#| eval: false

^.*\.Rproj$
^\.Rproj\.user$
```

</details>

### 3. "R" Folder

Within your package folder, make a new folder called "R". All of the functions in your package _must_ live in this folder but for now we can leave it empty.

:::

## Add a License

While R packages _tend_ to be publicly available, you still need to specify how you want people to be able to use the package and how to credit their use of your package! A "LICENSE" file is how this is specified. Information about choosing a `LICENSE` is provided in the R Package (2e) book [Chapter 12: Licensing](https://r-pkgs.org/license.html).

We'll use the Apache 2.0 license for now and use one of the license-specific functions from the `usethis` package to quickly build it into our nascent R package. Copy and run the code from the following chunk.

This will do three necessary things:

1. Create a LICENSE.md file with the legal text of the license
2. Add the name and version of the license to the relevant line of your DESCRIPTION file
3. Add LICENSE.md to your .Rbuildignore

```{r}
#| label: add-license
#| eval: false

usethis::use_apache_license()
```

```         
✔ Setting License field in DESCRIPTION to 'Apache License (>= 2.0)'
✔ Writing 'LICENSE.md'
✔ Adding '^LICENSE\\.md$' to '.Rbuildignore'
```

## Add Code

We are now ready to actually add code to our package! Go ahead and make a new R script in your "R" folder and name it "cv.R". Once you have that file, **click the collapsed menu below and copy/paste all of its contents into your empty "cv.R" file.**

<details> 
<summary>`cv()` Function - Version 1</summary>

```{r}
#| label: cv-fxn_version-01
cv <- function(x){

  # Calculate SD and mean
  sd_x <- sd(x = x, na.rm = FALSE)
  avg_x <- mean(x = x, na.rm = FALSE)

  # Calculate coefficient of variation
  coef_var <- sd_x / avg_x

  # Return that 
  return(coef_var)
}
```

</details>

## Add Documentation

Documentation is _vital_ for functions, particularly those shared as part of a package, because users will encounter errors and will want to refer back to some sort of help document to try to figure out where they went wrong. The preferred method of documenting a function is to use `roxygen2` syntax.

Work through the tabs below to create documentation in your function and get your package to recognize it as such.

:::{.panel-tabset}
### 1. Document the Function

**Add the following lines to the top of your `cv()` function in your "cv.R" file** (i.e., push down the 'actual' function code so that all documentation can be above the function itself).

```{r}
#| label: cv-fxn-docs
#' @title Calculate CV
#'
#' @description Calculates the Coefficient of Variation ("CV") of a vector of numbers.
#' 
#' @param x Vector of numbers
#'
#' @returns Coefficient of variation of the numbers provided to 'x'
#' @export
#' 
#' @examples
#' # Calculate CV
#' cv(x = c(4, 5, 6))
#' 
```

Double check that you're on the right track by confirming that your "cv.R" file looks like the code chunk in the collapsed menu below

<details> 
<summary>`cv()` Function - Version 2</summary>

```{r}
#| label: cv-fxn_version-02
#' @title Calculate CV
#'
#' @description Calculates the Coefficient of Variation ("CV") of a vector of numbers.
#' 
#' @param x Vector of numbers
#'
#' @returns Coefficient of variation of the numbers provided to 'x'
#' @export
#' 
#' @examples
#' # Calculate CV
#' cv(x = c(4, 5, 6))
#' 
cv <- function(x){

  # Calculate SD and mean
  sd_x <- sd(x = x, na.rm = FALSE)
  avg_x <- mean(x = x, na.rm = FALSE)

  # Calculate coefficient of variation
  coef_var <- sd_x / avg_x

  # Return that 
  return(coef_var)
}
```

</details>

### 2. Document the Package

Once your `cv()` function has `roxygen2`-style comments, we can document the _package_. Fortunately, the `devtools` package has a nice function for doing exactly this! **Run the following code chunk in your Console.**

```{r}
#| label: devtools-document
#| eval: false

devtools::document()
```

This will create a new folder (named "man", short for "manuals") and fill it with specially-formatted ".Rd" files. _You will **never** edit these files manually!_ If you update your function documentation and need to update the respective .Rd files, re-run `devtools::document()`!

### 3. Celebrate!

Congratulations, you're now the proud owner of your very own R package! You can now install this package (from GitHub) as well as run automatic checks on it, and even send it to CRAN.

A smaller, but only slightly less significant, win is that you can now check the help file for your `cv()` function! **Run the following code chunk in your Console.**

```{r}
#| label: check-help
#| eval: false

?cv
```

This should open up the help pane of your IDE and present the documentation that you just added (but processed so that it looks almost like a Quarto/RMarkdown document).

:::

## Testing

Now that our package has a function, we can start creating "unit tests" to make sure our function(s) behave as expected/as they should. The `testthat` package provides some really nice tools for doing this reproducibly and to a high standard so we'll use that structure.

:::{.panel-tabset}
### 1. Initialize `testthat`

Set up your package to use the `testthat` package by running the following code chunk in the Console.

```{r}
#| label: setup-testthat
#| eval: false

usethis::use_testthat()
```

This will create a "tests" folder that contains both an R script called "testthat.R"--which you can ignore--and a folder called "testthat" containing nothing (yet).

### 2. Write a Unit Test

That done, we can write specific unit tests for the `cv()` function that use functions from `testthat`. The syntax of a unit test with `testthat` can be a little alien until you get familiar with it so for now, **do the following:**

1. Make a new script called "test-cv.R" in the "testthat" folder inside of the "tests" folder
    - So the file path will be `pkg_{lastname} / tests / testthat / test-cv.R`
2. Click the collapsed menu below and copy/paste all of its contents into the "test-cv.R" file

<details> 
<summary>`cv()` Unit Test</summary>

```{r}
#| label: cv-unittest
#| eval: false

# Run all tests in this script:
## testthat::test_file(file.path("tests", "testthat", "test-cv.R"))

# Error testing
# test_that("Errors work as desired", {
#   # None in this function (yet)
#   expect_error()
# })

# Warning testing
test_that("Warnings work as desired", {
   expect_warning(cv(x = NULL))
})

# Message testing
# test_that("Messages work as desired", {
#   # None in this function (yet)
#   expect_message()
# })

# Output testing
test_that("Outputs are as expected", {
  
  # Make a testing vector
  test_vec <- c(4, 5, 6)
  
  # Use the function to calculate CV and calculate it by hand
  fxn_cv <- cv(x = test_vec)
  raw_cv <- (sd(x = test_vec, na.rm = FALSE) / mean(x = test_vec, na.rm = FALSE))

  # Check certain aspects of output
  expect_equal(fxn_cv, raw_cv)
  expect_true(class(fxn_cv) == "numeric")
})
```

</details> 

### 3. Test the Package

Now we can run all of the unit tests in our package--currently just the one--by running the following code chunk in the Console.

```{r}
#| label: devtools-test
#| eval: false

devtools::test()
```

This will return a handy little report--in the Console--summarizing the results of your unit tests.

:::

### Checking and Installing

Now that you've completed testing your package, you can check it for consistency and completeness using `devtools::check()`.

```{r}
#| eval: false

devtools::check()
```

```         
══ Documenting ════════════════════════════════════════════════════════════════════════════════════════════════════════════════════
ℹ Updating mytools documentation
ℹ Loading mytools

══ Building ═══════════════════════════════════════════════════════════════════════════════════════════════════════════════════════
Setting env vars:
• CFLAGS    : -Wall -pedantic -fdiagnostics-color=always
• CXXFLAGS  : -Wall -pedantic -fdiagnostics-color=always
• CXX11FLAGS: -Wall -pedantic -fdiagnostics-color=always
• CXX14FLAGS: -Wall -pedantic -fdiagnostics-color=always
• CXX17FLAGS: -Wall -pedantic -fdiagnostics-color=always
• CXX20FLAGS: -Wall -pedantic -fdiagnostics-color=always
── R CMD build ────────────────────────────────────────────────────────────────────────────────────────────────────────────────────
✔  checking for file ‘/home/dolinh/mytools/DESCRIPTION’ (610ms)
─  preparing ‘mytools’:
✔  checking DESCRIPTION meta-information (338ms)
─  checking for LF line-endings in source and make files and shell scripts
─  checking for empty or unneeded directories
─  building ‘mytools_0.0.0.9000.tar.gz’
   
══ Checking ═══════════════════════════════════════════════════════════════════════════════════════════════════════════════════════
Setting env vars:
• _R_CHECK_CRAN_INCOMING_REMOTE_               : FALSE
• _R_CHECK_CRAN_INCOMING_                      : FALSE
• _R_CHECK_FORCE_SUGGESTS_                     : FALSE
• _R_CHECK_PACKAGES_USED_IGNORE_UNUSED_IMPORTS_: FALSE
• NOT_CRAN                                     : true
── R CMD check ────────────────────────────────────────────────────────────────────────────────────────────────────────────────────
─  using log directory ‘/tmp/Rtmp1UgqFD/file6d79323df6fae/mytools.Rcheck’ (649ms)
─  using R version 4.2.2 (2022-10-31)
─  using platform: x86_64-pc-linux-gnu (64-bit)
─  using session charset: UTF-8
─  using options ‘--no-manual --as-cran’
✔  checking for file ‘mytools/DESCRIPTION’
─  this is package ‘mytools’ version ‘0.0.0.9000’
─  package encoding: UTF-8
✔  checking package namespace information
✔  checking package dependencies (2.1s)
✔  checking if this is a source package
✔  checking if there is a namespace
✔  checking for executable files
✔  checking for hidden files and directories
✔  checking for portable file names
✔  checking for sufficient/correct file permissions
✔  checking serialization versions
✔  checking whether package ‘mytools’ can be installed (3.2s)
✔  checking installed package size
✔  checking package directory
✔  checking for future file timestamps (412ms)
✔  checking DESCRIPTION meta-information (584ms)
✔  checking top-level files ...
✔  checking for left-over files
✔  checking index information
✔  checking package subdirectories ...
✔  checking R files for non-ASCII characters ...
✔  checking R files for syntax errors ...
✔  checking whether the package can be loaded (481ms)
✔  checking whether the package can be loaded with stated dependencies ...
✔  checking whether the package can be unloaded cleanly ...
✔  checking whether the namespace can be loaded with stated dependencies ...
✔  checking whether the namespace can be unloaded cleanly (450ms)
✔  checking loading without being on the library search path (522ms)
✔  checking dependencies in R code (1.2s)
✔  checking S3 generic/method consistency (1s)
✔  checking replacement functions ...
✔  checking foreign function calls ...
✔  checking R code for possible problems (5.2s)
✔  checking Rd files (449ms)
✔  checking Rd metadata ...
✔  checking Rd line widths ...
✔  checking Rd cross-references ...
✔  checking for missing documentation entries ...
✔  checking for code/documentation mismatches (885ms)
✔  checking Rd \usage sections (1.3s)
✔  checking Rd contents ...
✔  checking for unstated dependencies in examples ...
✔  checking examples (2.7s)
✔  checking for unstated dependencies in ‘tests’ ...
─  checking tests (418ms)
✔  Running ‘testthat.R’ (1.4s)
✔  checking for non-standard things in the check directory
✔  checking for detritus in the temp directory
   
   
── R CMD check results ──────────────────────────────────────────────────────────────────────────────────── mytools 0.0.0.9000 ────
Duration: 27.3s

0 errors ✔ | 0 warnings ✔ | 0 notes ✔
```

Then you can install it locally using `devtools::install()`, which needs to be run from the parent directory of your module

```{r}
#| eval: false

devtools::install()
```

```         
── R CMD build ────────────────────────────────────────────────────────────────────────────────────────────────────────────────────
✔  checking for file ‘/home/dolinh/mytools/DESCRIPTION’ (541ms)
─  preparing ‘mytools’:
✔  checking DESCRIPTION meta-information ...
─  checking for LF line-endings in source and make files and shell scripts
─  checking for empty or unneeded directories
─  building ‘mytools_0.0.0.9000.tar.gz’
   
Running /opt/R/4.2.2/lib/R/bin/R CMD INSTALL /tmp/Rtmp1UgqFD/mytools_0.0.0.9000.tar.gz --install-tests 
* installing to library ‘/home/dolinh/R/x86_64-pc-linux-gnu-library/4.2’
* installing *source* package ‘mytools’ ...
** using staged installation
** R
** tests
** byte-compile and prepare package for lazy loading
** help
*** installing help indices
** building package indices
** testing if installed package can be loaded from temporary location
** testing if installed package can be loaded from final location
** testing if installed package keeps a record of temporary installation path
* DONE (mytools)
```

After installing, your package is now available for use in your local environment, yay!

::: callout-important
##### Check out the Build Tab

Remember when we ran `usethis::create_package()` and after we ran it we saw the Build Tab added to the Environment pane?

In the Build Tab, each of the buttons correspond with one of the `devtools` functions we ran, meaning:

-   Test button is equivalent to running `devtools::test()` in the Console
-   Check button is equivalent to running `devtools::check()` in the Console
-   Install button is equivalent to running `devtools::install()` in the Console
:::

### Sharing and Releasing

-   **GitHub**: The simplest way to share your package with others is to upload it to a [GitHub repository](https://github.com), which allows others to install your package using the `install_github('mytools','github_username')` function from `devtools`.

-   **CRAN**: If your package might be broadly useful, also consider releasing it to [CRAN](https://cran.r-project.org/), using the `release()` method from `devtools()`. Releasing a package to CRAN requires a significant amount of work to ensure it follows the standards set by the R community, but it is entirely tractable and a valuable contribution to the science community. If you are considering releasing a package more broadly, you may find that the supportive community at [ROpenSci](https://ropensci.org) provides incredible help and valuable feeback through their onboarding process.

-   **R-Universe**: A newer approach is to link your package release to [R-Universe](https://r-universe.dev), which is an effective way to make it easy to test and maintain packages so that many people can install them using the familiar `install.pacakges()` function in R. In R-Universe, people and organizations can [create their own *universe* of packages](https://github.com/r-universe-org/help#how-to-setup-your-personal-universe), which represent a collection of packages that appear as a CRAN-compatible repository in R. For example, for DataONE we maintain the [DataONE R-Universe](https://dataoneorg.r-universe.dev), which lists the packages we actively maintain as an organization. So, any R-user that wants to install these packages can do so by adding our universe to their list of repositories, and then installing packages as normal. For example, to install the `codyn` package, one could use:

```{r}
#| eval: false

install.packages('codyn', repos = c('https://dataoneorg.r-universe.dev', 'https://cloud.r-project.org'))
```

### Exercise: Add More Functions

Add additional temperature conversion functions to the `mytools` package and:

-   Add full documentation for each function
-   Write tests to ensure the functions work properly
-   Rebuild the package using `document()`, `check()`, and `install()`

::: callout-important
#### Don't forget to update the version number before you install!

Version information is located in the `DESCRIPTION` file and when you first create a package the version is 0.0.0.9000.

This version number follows the format major.minor.patch.dev. The different parts of the version represent different things:

-   **Major**: A significant change to the package that would be expected to break users code. This is updated very rarely when the package has been redesigned in some way.
-   **Minor**: A minor version update means that new functionality has been added to the package. It might be new functions to improvements to existing functions that are compatible with most existing code.
-   **Patch**: Patch updates are bug fixes. They solve existing issues but don't do anything new.
-   **Dev**: Dev versions are used during development and this part is missing from release versions. For example you might use a dev version when you give someone a beta version to test. A package with a dev version can be expected to change rapidly or have undiscovered issues.

After you've made some changes to a package, but before you install run the code:

```{r}
#| eval: false
usethis::use_version()
```

```         
Current version is 0.0.0.9000.
What should the new version be? (0 to exit) 

1: major --> 1.0.0
2: minor --> 0.1.0
3: patch --> 0.0.1
4:   dev --> 0.0.0.9001
```

Since we're adding new functions, we can consider this a *minor* change and can select option 2.

```         
Selection: 2
✔ Setting Version field in DESCRIPTION to '0.1.0'
```

*Source: COMBINE's R package workshop, [Ch 9: Versioning](https://combine-australia.github.io/r-pkg-dev/versioning.html)*
:::

### Add Dependencies

If your R code depends on functions from another package, you must declare it. In the `Imports` section in the `DESCRIPTION` file, list all the packages your functions depend upon.

In our `custom_theme()` function, we depend on the `ggplot2` package, and so we need to list it as a dependency.

Once again, `usethis` provides a handy helper method:

```{r}
#| eval: false

usethis::use_package("ggplot2")
```

```         
✔ Adding 'ggplot2' to Imports field in DESCRIPTION
• Refer to functions with `ggplot2::fun()`
```

Take a look at the `DESCRIPTION` file again, and you'll see the Imports section has been added, with `ggplot2` underneath.

::: {.callout-tip appearance="minimal"}
```         
Package: mytools
Title: Halina Do-Linh's Utility R Functions
Version: 0.0.0.9000
Authors@R: 
    person("Halina", "Do-Linh", email = "dolinh@nceas.ucsb.edu", role = c("aut", "cre"),
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: A collection of useful R functions that I use for general utilities.
License: Apache License (>= 2)
Encoding: UTF-8
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.2.3
Imports: 
    ggplot2
```
:::

### Additional Resources

-   Hadley Wickham and Jenny Bryan's awesome book: [R Packages](https://r-pkgs.org/)
-   ROpenSci Blog Post: [How to create your personal CRAN-like repository on R-universe](https://ropensci.org/blog/2021/06/22/setup-runiverse/)
-   Karl Broman's: [R package primer: a minimal tutorial on writing R packages](https://kbroman.org/pkg_primer)
-   Thomas Westlake's Short Tutorial: [Writing an R package from scratch](https://r-mageddon.netlify.com/post/writing-an-r-package-from-scratch/) (his post is an updated version of Hilary Parker's [blog post](https://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/))
